Hydra Battles and Cie

### 2021-08-09 doc/chapter-primrec.tex
fix typos in chapter-powers.tex and chapter-primrec.tex

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of scripts. The part dedicated to primitive recursive functions and Peano arithmetics is 32K lines long and is originally structured in 38 modules.
Thus, we propose a partial exploration of this library, through examples and exercises. Our additions to the original library --- mainly examples and counter-examples ---,
are stored in the directory \texttt{theories/ordinals/MoreAck}.

@@

After this test, the term \texttt{bigPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections simple ways to prove that a given function is primitive recursive, whithout building such an unreadable term.

@@

Thus, the proof that a function, like \texttt{factorial}, is primitive recursive may be interactive, whithout having to type complex terms at any step of the development.

@@

\index{primrec}{Predicates!isPR}
\index{coq}{Extensionnaly equal functions}

Let $f$ be an arithmetic function of arity $n$. We say that $f$ is primitive recursive if $f$ is \textbf{extensionnaly}
equal to the interpretation of some term of type \texttt{PrimRec $n$}.

@@

We prefered to define a variant which uses explicitely 
 the functional \texttt{iterate},
where (\texttt{iterate\,$f$\,$n$})
is the $n$-th iteration of $f$\,\footnote{Please not confuse with \texttt{primRec.iterate}, which is monomorphic and does not share the same order of arguments.}. It makes it possible to apply a few lemmas proved in 
\href{../theories/html/hydras.Prelude.Iterates.html}{Prelude.Iterates}, for instance about the monotony of the $n$-th iterate of a given function.

@@

\begin{exercise}
The file \href{../theories/html/hydras.MoreAck.Ack.html}{MoreAck.Ack} presents two other definitions of the Ackermann functions based on the lexicographic ordering on $\mathbb{N}\times\mathbb{N}$.
Prove that the four functions are extensionnally equal.
\end{exercise}

@@

The three first lemmas make us sure that our function 
\texttt{Ack} satifies the ``usual'' equations.

@@

Ìn order to prove that \texttt{Ack} (considered as a function of two arguments) is not primitive recursive, the usual method consists in two steps:

@@

Please note that, in order to prove a property shared by any primitive recursive function of, say, arity 2, this induction scheme  leads you to consider an expension of the considered property to primitive recursive function of any arity.

@@

\begin{remark}
It is easy to prove that any unary function which dominatates \texttt{fun n => Ack n n} fails to be primitive recursive. To this end, we use an instance of \texttt{majorAnyPR} dealing with unary functions.

@@

\begin{remark}
Note that the Ackermann function is a counter-example to the (false) statement:
\begin{quote}
{\color{red}
  ``Let $f$ be a function of type \texttt{naryFunc\,2}. If, for any $n$, the fonction $f(n)$ is primitive recursive, then f is primitive recursive.''}
\end{quote}
\end{remark}

@@

In order to get rid of the substraction in the definition of \texttt{l\_std}, we work with a helper function.

@@

Our proof of this lemma is not trivial at all, it uses some properties of the Ketonen-Solovay's toolkit. We advise the reader to explore this proof, with the help of an ide or software like alectryon.
%%
%%% To move the path chapter
%%%%

>>>>

O'Connor's library on Gödel's incompleteness theorems contains a little more 
than 45K lines of scripts. The part dedicated to primitive recursive functions and Peano arithmetic is 32K lines long and is originally structured in 38 modules.
Thus, we propose a partial exploration of this library, through examples and exercises. Our additions to the original library --- mainly examples and counter-examples ---,
are stored in the directory \texttt{theories/ordinals/MoreAck}.

@@

After this test, the term \texttt{bigPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections simple ways to prove that a given function is primitive recursive, without building such an unreadable term.

@@

Thus, the proof that a function, like \texttt{factorial}, is primitive recursive may be interactive, without having to type complex terms at any step of the development.

@@

\index{primrec}{Predicates!isPR}
\index{coq}{Extensionally equal functions}

Let $f$ be an arithmetic function of arity $n$. We say that $f$ is primitive recursive if $f$ is \textbf{extensionally}
equal to the interpretation of some term of type \texttt{PrimRec $n$}.

@@

We preferred to define a variant which uses explicitly
 the functional \texttt{iterate},
where (\texttt{iterate\,$f$\,$n$})
is the $n$-th iteration of $f$\,\footnote{Please not confuse with \texttt{primRec.iterate}, which is monomorphic and does not share the same order of arguments.}. It makes it possible to apply a few lemmas proved in 
\href{../theories/html/hydras.Prelude.Iterates.html}{Prelude.Iterates}, for instance about the monotony of the $n$-th iterate of a given function.

@@

\begin{exercise}
The file \href{../theories/html/hydras.MoreAck.Ack.html}{MoreAck.Ack} presents two other definitions of the Ackermann functions based on the lexicographic ordering on $\mathbb{N}\times\mathbb{N}$.
Prove that the four functions are extensionally equal.
\end{exercise}

@@

The three first lemmas make us sure that our function 
\texttt{Ack} satisfies the ``usual'' equations.

@@

In order to prove that \texttt{Ack} (considered as a function of two arguments) is not primitive recursive, the usual method consists in two steps:

@@

Please note that, in order to prove a property shared by any primitive recursive function of, say, arity 2, this induction scheme  leads you to consider an extension of the considered property to primitive recursive function of any arity.

@@

\begin{remark}
It is easy to prove that any unary function which dominates \texttt{fun n => Ack n n} fails to be primitive recursive. To this end, we use an instance of \texttt{majorAnyPR} dealing with unary functions.

@@

\begin{remark}
Note that the Ackermann function is a counter-example to the (false) statement:
\begin{quote}
{\color{red}
  ``Let $f$ be a function of type \texttt{naryFunc\,2}. If, for any $n$, the function $f(n)$ is primitive recursive, then f is primitive recursive.''}
\end{quote}
\end{remark}

@@

In order to get rid of the subtraction in the definition of \texttt{l\_std}, we work with a helper function.

@@

Our proof of this lemma is not trivial at all, it uses some properties of the Ketonen-Solovay's toolkit. We advise the reader to explore this proof, with the help of an IDE or software like Alectryon.
%%
%%% To move the path chapter
%%%%
